---
title: "Learn R From Scratch"
format: html
editor: visual
copyright: "Barthélémy DURETTE 2024 all rights reserved"
---

------------------------------------------------------------------------

# Manipulating Data with R

------------------------------------------------------------------------

### Welcome to this third module of the "Learn R From Scratch" training course.

In this module, we will cover:

1.  **How to select rows from a data table**

2.  **How to select columns from a data table**

3.  **How to add new variables to a data table**

In this module, you will also learn how to use the **Tidyverse package and the pipe operator**: **%\>%**

------------------------------------------------------------------------

## Introduction: The Tidyverse Package

The Tidyverse library marked a turning point in the history of R. This package, which actually combines several libraries designed to work together, changed the way people program in R. It shifted programming from a mainly functional style to a pipeline-based approach. It also introduced a set of highly useful functions for data manipulation. We will therefore begin by installing it. This may take a few minutes.

In addition, the ggplot2 package is required.

```{r}
install.packages("tidyverse")
library(tidyverse)
library(ggplot2)
```

To practice data manipulation, we will use a dataset containing more than 50,000 rows, which would be difficult to manage in a traditional spreadsheet.

This is the **diamonds** dataset, whose characteristics can be explored as follows:

```{r}
help("diamonds")
```

## 1 - Selecting Rows from a Data Table

This operation is also called **filtering**, because it consists of keeping only the rows that satisfy a logical condition. For example, let us consider the quality of the diamond cut: the `cut` variable.

The variable **cut** is a **factor** that classifies diamonds into five categories:

```{r}
levels(diamonds$cut)
```

We can count the number of diamonds in each category using the **table()** function.

```{r}
table(diamonds$cut)
```

To filter a dataset, we use the **filter()** function. The argument of this function is the logical condition to apply. For example:

- **cut == "Fair"** keeps only diamonds with a Fair cut quality.

```{r}
# Restrict the dataset to diamonds with a Fair cut

diamonds %>% filter(cut == "Fair", color == "E")
```

- **cut != "Fair"** keeps only diamonds whose cut quality is not Fair.

```{r}
# Restrict the dataset to diamonds whose cut is not Fair

diamonds %>% filter(cut != "Fair")
```

- **cut %in% c("Fair","Good")** keeps all diamonds in the Fair or Good categories.

```{r}
# Restrict the dataset to diamonds with Fair or Good cuts

choices <- c("Fair","Good","Ideal")
diamonds %>% filter(cut %in% choices)
```

**Along the way, you have encountered:**

- the **pipe operator %\>%**

- the **%in% operator**

- the **c() function**

------------------------------------------------------------------------

### Practice Exercise

```         
In your own words, describe what the following three R programming elements do:

The %>% operator passes the output of one function to the input of another function.

The %in% operator is a logical connector that tests membership in a set of elements.

The c() function creates a vector from the elements provided as arguments.
```

------------------------------------------------------------------------

### Application Exercise

A filter can be applied to any variable, even if it is not a factor. For example, we can filter a numerical variable.

Filter the diamonds dataset to keep only diamonds worth more than \$15,000.

```{r}
# Write your code here:

diamonds %>% filter(price > 15000)
```

**How many diamonds have a price greater than \$15,000?**

------------------------------------------------------------------------

## 2 - Selecting Columns from a Data Table

The Tidyverse package also provides a very useful function for selecting columns: **select()**.

This function can be used to select:

- One or more columns **by name**

- All columns **except one or more**

- All columns **starting or ending with the same prefix or suffix**

- All columns **containing a specific character string**

- etc. (see the function documentation for all available options)

Here are some examples:

### Selecting a subset of columns by name

```{r}
# Select the Cut, Clarity, and Price columns

diamonds %>% select(cut, clarity, price)

# Note: column names are not enclosed in quotation marks
```

### Selecting all columns except specific ones

```{r}
# Exclude the Cut, Clarity, and Price columns using the - operator

diamonds %>% select(-cut, -clarity, -price)

# Exclude the same columns using ! and c()

diamonds %>% select(!c(cut, clarity, price))
```

### Selecting all columns that start with a given string

```{r}
# Select all columns whose names start with the letter "c"

diamonds %>% select(starts_with("c"))
```

------------------------------------------------------------------------

### Application Exercise

Using the diamonds dataset, create a table containing only the dimensions (`x`, `y`, and `z`) and the price of diamonds whose cut quality is Ideal.

```{r}
# Write your code here:

```

**What are the dimensions of this new dataset?**

------------------------------------------------------------------------

## 3 - Adding New Variables to a Data Table

In data analysis, it is common to compute intermediate variables, often called **features** in data mining. This is especially useful when several variables need to be combined to create a value that is more meaningful for analysis.

For example, the diamonds dataset provides the **size of diamonds in three spatial dimensions** (`x`, `y`, and `z`). Individually, these variables are of limited interest. However, they can be combined to obtain an **estimate of diamond volume**.

Our goal here is to calculate the estimated volume of each diamond and **add this value to the dataset** so that it can be used in later analyses. We use the **mutate()** function for this purpose.

```{r}
# Add a volume column to the diamonds dataset

diamonds %>% mutate(volume = x * y * z)
```

We can then use this new column in subsequent operations. For example, we can display diamond weight as a function of estimated volume.

```{r}
# Scatter plot of diamond weight as a function of volume

diamonds %>%
  mutate(volume = x * y * z) %>%
  ggplot() +
  geom_point(aes(x = volume, y = carat))
```

**Does anything surprise you about the resulting graph? What do you think explains it?**

------------------------------------------------------------------------

### Application Exercise

Using the diamonds dataset:

1.  Calculate the estimated volume of each diamond from its dimensions `x`, `y`, and `z`.

2.  Filter out diamonds with a volume smaller than 10 mm³ or larger than 1000 mm³.

3.  Visualize diamond price (`price`) as a function of volume.

4.  Map the color of the points to diamond clarity (`clarity`).

```{r}
# Write your code here:

```

**Which level of diamond clarity appears to be most valued by buyers?**

Propose a clearer visualization to illustrate this trend using the histogram geometry introduced in the previous module.

```{r}
# Write your code here:

```
