---
title: "Learn R From Scratch"
author: "Barthélémy Durette 2025 all rights reserved"
format: html
editor: visual
---

## Loading Your Own Data Files

------------------------------------------------------------------------

### Welcome to this fifth module of the "Learn R From Scratch" training course.

In this module, we will cover:

1.  **How to load a .csv data file into R?**

2.  **How to load an Excel data file into R?**

To do so, we will use the **tidyverse** package and the **readxl** package.

```{r}
require(tidyverse)
require(readxl)
```

------------------------------------------------------------------------

## 1 - Loading a .csv Data File

The **.csv** file format (Comma Separated Values) is an extremely common way of storing **tabular data**. In this format, which can be opened with a simple text editor, **data are stored row by row as values separated by commas** (or semicolons). The first row, called the **header**, contains the variable names, in other words the column names.

**Example: the first 10 rows of the file snore.csv**

```         
"","AGE","POIDS","TAILLE","ALCOOL","SEXE","RONFLE","TABA","PDS","agestd","alcool","alcool2"
"1","47","71","158","0","1","N","O","0.6667","-0.462394463139037","0","0"
"2","56","58","164","7","1","O","N","0.6667","0.327273500476016","3","3"
"3","46","116","208","3","1","N","O","0.6667","-0.550135347985154","1","1"
"4","70","96","186","3","1","N","O","0.6667","1.55564588832165","1","1"
"5","51","91","195","2","1","O","O","0.6667","-0.111430923754569","1","1"
"6","46","98","188","0","2","N","N","2","-0.550135347985154","0","0"
"7","40","112","193","5","1","N","O","0.6667","-1.07658065706186","2","2"
"8","46","77","165","0","2","N","N","2","-0.550135347985154","0","0"
"9","49","76","164","0","1","N","N","0.6667","-0.286912693446803","0","0"
```

***Which other character is used here to separate values? Why?***

To read a .csv file, you can use the **read.csv()** function available in base R or the **read_csv()** function available in the tidyverse package and compatible with the pipe operator.

Import using read.csv():

```{r}
# Import the file snore.csv using read.csv()
snore <- read.csv("../datasets/snore.csv")

# Preview the file
head(snore)
```

Import using read_csv():

```{r}
# Import the file snore.csv using read_csv()
snore_tidy <- read_csv("../datasets/snore.csv")

# Preview the file
head(snore_tidy)
```

*What differences do you notice between the output of **read.csv()** and **read_csv()**?*

**The function parameters allow you to adapt to the specific characteristics of your .csv file.** For example, let us replace commas with semicolons in the file "snore.csv" to create the file "snore_semi.csv".

**First 10 rows of the file snore_semi.csv:**

```         
"";"AGE";"POIDS";"TAILLE";"ALCOOL";"SEXE";"RONFLE";"TABA";"PDS";"agestd";"alcool";"alcool2"
"1";"47";"71";"158";"0";"1";"N";"O";"0.6667";"-0.462394463139037";"0";"0"
"2";"56";"58";"164";"7";"1";"O";"N";"0.6667";"0.327273500476016";"3";"3"
"3";"46";"116";"208";"3";"1";"N";"O";"0.6667";"-0.550135347985154";"1";"1"
"4";"70";"96";"186";"3";"1";"N";"O";"0.6667";"1.55564588832165";"1";"1"
"5";"51";"91";"195";"2";"1";"O";"O";"0.6667";"-0.111430923754569";"1";"1"
"6";"46";"98";"188";"0";"2";"N";"N";"2";"-0.550135347985154";"0";"0"
"7";"40";"112";"193";"5";"1";"N";"O";"0.6667";"-1.07658065706186";"2";"2"
"8";"46";"77";"165";"0";"2";"N";"N";"2";"-0.550135347985154";"0";"0"
"9";"49";"76";"164";"0";"1";"N";"N";"0.6667";"-0.286912693446803";"0";"0"
```

To read this new file, you must specify in the function parameters that the separator between values is a semicolon:

```{r}
# Import snore_semi.csv using read.csv()
snore_semi <- read.csv("../datasets/snore_semi.csv",
                      sep = ";")

# Import snore_semi.csv using tidyverse
snore_semi_tidy <- read_delim("../datasets/snore_semi.csv",
                             delim = ";")
```

## 2 - Loading an Excel Data File

To load data from an Excel file, we will use the **read_xlsx()** function available in the readxl package. Unlike .csv files, Excel files have a standardized format. However, it is possible to choose which worksheet should be imported.

```{r}
# Read the file "snore.xlsx" using read_xlsx()

snore_xlsx <- read_xlsx("../datasets/snore.xlsx",
                         sheet = 1)
```

### *Practice Exercise*

Working with a partner, load your own data from a CSV or Excel file:

```{r}
# Write your code here:

```
