---
title: "Learn R From Scratch"
format: pdf
editor: visual
copyright: "Barthélémy DURETTE 2024 all rights reserved"
---

------------------------------------------------------------------------

# Introduction

------------------------------------------------------------------------

### Welcome to this first module of the training course "Learn R From Scratch".

In this module, we will cover:

1.  **How to perform basic arithmetic operations with R?**

2.  **How to view a data table in R?**

3.  **How to perform basic arithmetic operations on a data table?**

You will learn how to use **\$**, the first essential symbol you need to know when working with R.

------------------------------------------------------------------------

## 1 - Performing Basic Arithmetic Operations with R

At a minimum, R can be used as a powerful calculator. Addition, subtraction, multiplication, and division are written as follows (click the green arrow in the upper-left corner of each cell to display the result):

Addition

```{r}
# Addition
1+1
```

Subtraction

```{r}
# Subtraction
1-1
```

Multiplication

```{r}
# Multiplication
2*2
```

Division

```{r}
# Division
4/2
```

For exponentiation, use the **\^** symbol.

```{r}
# Power
2^3
```

For square roots, you can either use the **sqrt** function or raise a number to the power of 1/2.

```{r}
# Square root using sqrt
sqrt(16)

# Square root using exponentiation
16^(1/2)
```

R also provides all standard trigonometric functions:

```{r}
# Cosine
cos(20)

# Sine
sin(10)

# Tangent
tan(10)

# Arctangent
atan(10)
```

And logarithmic functions:

```{r}
# Natural logarithm (base e)
log(1)

# Base-10 logarithm using log10
log10(10)

# Base-10 logarithm using log
log(10, base = 10)
```

That concludes a first overview of the most commonly used functions.

------------------------------------------------------------------------

### *Practice Exercise*

Calculate the base-10 logarithm of the sine of 8 raised to the power of 3.

```{r}
# Write your formula below this line:

```

Do you get **-1.099532**? Great! Otherwise, look for the mistake.

------------------------------------------------------------------------

## 2 - Viewing a Data Table in R

R includes several built-in datasets that are ready to use. Some of the most popular are:

1.  mtcars: data on cars

2.  iris: data on flowers

3.  ToothGrowth: data on tooth growth

4.  PlantGrowth: data on plant growth

5.  USArrests: data on crime statistics

To learn more about these datasets, you can use the **help** function, which opens documentation in a separate RStudio tab. The **data()** function provides access to all built-in datasets.

```{r}
help("ToothGrowth")
```

To access a dataset, simply type its name. However, if you do that (go ahead and try!), R will display the entire dataset by default. That can be difficult to read. Here are a few useful functions for exploring a dataset.

The **dim** function returns the dimensions of the table. Here, there are 150 rows and 5 columns.

```{r}
# dim function
dim(iris)
```

The **names** function returns the column names:

```{r}
# names function
names(iris)
```

Finally, the **head** function provides a preview of the first rows of the dataset:

```{r}
head(iris)
```

------------------------------------------------------------------------

### *Practice Exercise*

Choose one of R's built-in datasets and:

- Consult the documentation to obtain general information about the dataset.
- Find the number of rows and columns.
- Find the column names.

```{r}
# Write your code here:

```

------------------------------------------------------------------------

## 3 - How to Perform Basic Arithmetic Operations on a Data Table?

Now let's get to the heart of the matter. How can you use R to manipulate data and perform basic operations?

|  |
|------------------------------------------------------------------------|
| **The first of the four essential symbols you need to know in R is the "\$" symbol (money, money, money!).** |

The **\$** symbol is used to select a column from a data frame. For example, if I want to retrieve the "Sepal.Length" column from the "iris" dataset, I would write:

```{r}
# Retrieve the Sepal.Length column from the iris dataset
iris$Sepal.Length
```

------------------------------------------------------------------------

### *Practice Exercise*

Calculate the length of `iris$Sepal.Length` and display only the first 5 values. Caution : dim() may not be the appropriate function to get the length of a single column, try an educated guess !

```{r}
# Write your answer here:

```

------------------------------------------------------------------------

You can then perform **arithmetic operations directly on columns**, such as:

Multiplying by a scalar:

```{r}
# Multiply the Petal.Length column by four
4 * iris$Petal.Length
```

Computing the sine:

```{r}
# Compute the sine of the Petal.Length column
sin(iris$Petal.Length)
```

You can also perform operations between columns, for example **adding two columns**:

```{r}
# Sum of the Petal.Length and Sepal.Length columns
iris$Petal.Length + iris$Sepal.Length
```

Or calculating the **element-wise product** of two columns:

```{r}
# Product of the Petal.Length and Sepal.Length columns
iris$Petal.Length * iris$Sepal.Length
```

------------------------------------------------------------------------

### *Practice Exercise*

Calculate the sine of the sum of the `Petal.Length` and `Sepal.Length` columns from the `iris` dataset.

```{r}
# Write your answer here:

```
